Cybersecurity Impacts of the Covid-19 Pandemic in Italy

### 2021-03-09 article.tex
Fixed typos

\begin{abstract}
  The Covid-19 pandemic has pushed companies to the extensive
  use of digital services, to implement home working and provide
  online services to people in lockdown. As a consequence, it is interesting
  to study how this has affected the number, kind and distribution of cybersecurity
  attacks. This paper gives an empirical evaluation of the cybersecurity attacks
  at the beginning of the Covid-19 pandemic in Italy, based on data collected from
  the questionnaires of the annual Digital Attacks Observatory. It shows that the overall
  number of attacks has not increased, but attacks have affected smaller companies
  than before. This can be explained with the fact that the Italian
  industrial scenario is mostly populated by small and medium enterprises,
  that have been obliged to a quick reconvertion of their IT systems and typically
  lack the necessary cybersecurity culture.

@@

About half of the OAD questionnaires refer to the occurred
cyberattack typologies and their main characteristics.
The other half refers to the technical and organizational security measures implemented
on the IT systems of the respondents. 
Therefeore, OAD allows one to match the cyberattacks to the cyber-measures in each production
IT system considered. 
In fact, OAD provides a qualitative macro-evaluation of the implemented measures
(within the specific context of the company), in order to motivate the anonymous respondents
to complete the questionnaires.
In particular, the security measures considered in the 2020 questionnaire were subdivided
in \emph{technical}, \emph{organizational}, \emph{managerial} and \emph{governamental}, as follow:

\begin{enumerate}
\small
	\item \emph{Technical measures}
	\begin{itemize}
		\item Overall architecture of the digital security measures, integrated with the entire IT system architecture
		\item Physical countermeasures
		\item Identification, Authentication, and Authorization (IAA) measures
		\item Cybersecurity measures at local and geographical network level
		\item Measures for logical protection of each single Information and Communication Technology (ICT) system
		\item Application and software protection
		\item Data protection
	\end{itemize}
	\item \emph{Organizational measures}
	\begin{itemize}
		\item Organizational structure, roles, skills, certifications
		\item Organizational policies and procedures
		\item Contracts, agreements and digital security clauses with third parties
		\item Awareness and culture of digital security at all levels
		\item Auditing
	\end{itemize}
	\item \emph{Managerial and governamental measures}
	\begin{itemize}
		\item Digital security control, monitoring and management systems
		\item Disaster Recovery (DR) plan.
	\end{itemize}
\end{enumerate}

@@

\begin{figure}
  \centering
  \includegraphics[width=1\textwidth]{pictures/fig2.png}
  \caption{Trend of the reported attacks from the beginning of the OAD survey
    in 2008~\cite{oad20}. This comparison is only valid as a trend and not as a statistical
    measure, since the respondents are different from year to year. The dotted red line shows
    successive waves: after a relative peak of attacks, companies react and strengthen
    their cybersecurity measures, so that, the following year, the trend dicreases.}
  \label{fig:2}
\end{figure}

@@

The chart in Fig.~\ref{fig:2} analyzes the trend of cyberattaks.
The figure shows that, from 2007 to 2015, in average, about 40\% of the respondents
reported cyberattacks (leftmost full red line). Since 2015, the trend increased to 45\%
of the respondents (rightmost full red line). 
In 2018, OAD reached the peak of reported attacks and, for the first time, the percent of
reported attacks became larger than that percent of companies that did not report any attack.
After the peak 
in 2018, year 2019 has a lower figure and the same occurs in the first quarter of 2020.
The trend for full 2020 will be discovered with the currently ongoing OAD survey.
This view, over several years, does not show significant changes, not even inside
the pandemic period, but shows only a slight increase in cyberattacks 
in the recent years, not necessarily attributable to the pandemic crisis. One could wonder
why the percentages in Fig.~\ref{fig:2} are lower than what the reader would have expected, 
given the numerous cyber attacks occured in Italy. As already discussed at the beginning of 
this section, the main reason seems the size of the Italian companies.

@@

As pointed out in Fig.~\ref{fig:1}, the IT systems considered in the 2020 survey and their cybersecurity are in the medium to high range and the collected information shows a relevant 
improvement in both technical and organizational measures, in comparison to the previous surveys. Few of the considered Italian IT systems are based on data centers in 
Italy: most of the respondents' companies have medium to small IT systems partly on premise and partly outsourced, with an increasing use of cloud services.
The high number of attacks detected in 2018, and the privacy obligations related to the European
General Data Protection Regulation (GDPR), have certainly contributed to strengthening cybersecurity measures, and a further improvement 
comes from the growing use of cloud services, where high standard security measures are in place, usually. Organizational measures for cybersecurity, historically missing and neglected
in Italy, have improved in terms of defition of roles and separation of duties, of organizational policies and procedures, and of incident management. These measures often are missing
in small organizations, and in general the cybersecurity awareness and competence are low, and mainly concentrated at the top level of the public and private organizations. In Italy, there is 
still a long way to go in terms of cybersecurity continued training and awareness. A formal and bureaucratic approach to the organizational procedures is still prevalent. Once defined,
such procedures often do not find concrete application and misses periodic updates and operational tests. A significant example is the disaster recovery plans, for 
which, typically, the required alternative IT resources are neither forecast nor allocated. Periodic exercises and simulations of the 
possible disasters are not implemented. Cybersecurity management tools have limited diffusion yet, among the 2020 OAD respondents. This is particularly true for
the most advanced ones, based on artificial intelligence.
Although the use of IoT~\cite{JK21}, industrial automation, robotics systems~\cite{SGLMDXHKCZT21} and systems based on blockchain technology~\cite{RMFF21}
are expanding and can contribute to fight the Covid-19 pandemic.
In the OAD 2020 report, one finds only a few respondents involved in such technologies, which also derives from the limited portion
of their economic sector, interested in such systems: the manufacturing sectors, the logistics, the research and 
development centers and labs, the local public administration for territorial control, and so on.
%For these topics, the data that emerged from the survey are considered only as 
%specific cases that contribute to the general values on the cyberattacks, but which cannot be still considered of reference at the Italian level.

>>>>

\begin{abstract}
  The Covid-19 pandemic has pushed companies to the extensive
  use of digital services, to implement home working and provide
  online services to people in lockdown. As a consequence, it is interesting
  to study how this has affected the number, kind and distribution of cybersecurity
  attacks. This paper gives an empirical evaluation of the cybersecurity attacks
  at the beginning of the Covid-19 pandemic in Italy, based on data collected from
  the questionnaires of the annual Digital Attacks Observatory. It shows that the overall
  number of attacks has not increased, but attacks have affected smaller companies
  than before. This can be explained with the fact that the Italian
  industrial scenario is mostly populated by small and medium enterprises,
  that have been obliged to a quick reconversion of their IT systems and typically
  lack the necessary cybersecurity culture.

@@

About half of the OAD questionnaires refer to the occurred
cyberattack typologies and their main characteristics.
The other half refers to the technical and organizational security measures implemented
on the IT systems of the respondents. 
Therefore, OAD allows one to match the cyberattacks to the cyber-measures in each production
IT system considered. 
In fact, OAD provides a qualitative macro-evaluation of the implemented measures
(within the specific context of the company), in order to motivate the anonymous respondents
to complete the questionnaires.
In particular, the security measures considered in the 2020 questionnaire were subdivided
in \emph{technical}, \emph{organizational}, \emph{managerial} and \emph{governamental}, as follow:

\begin{enumerate}
\small
	\item \emph{Technical measures}
	\begin{itemize}
		\item Overall architecture of the digital security measures, integrated with the entire IT system architecture
		\item Physical countermeasures
		\item Identification, Authentication, and Authorization (IAA) measures
		\item Cybersecurity measures at local and geographical network level
		\item Measures for logical protection of each single Information and Communication Technology (ICT) system
		\item Application and software protection
		\item Data protection
	\end{itemize}
	\item \emph{Organizational measures}
	\begin{itemize}
		\item Organizational structure, roles, skills, certifications
		\item Organizational policies and procedures
		\item Contracts, agreements and digital security clauses with third parties
		\item Awareness and culture of digital security at all levels
		\item Auditing
	\end{itemize}
	\item \emph{Managerial and governmental measures}
	\begin{itemize}
		\item Digital security control, monitoring and management systems
		\item Disaster Recovery (DR) plan.
	\end{itemize}
\end{enumerate}

@@

\begin{figure}
  \centering
  \includegraphics[width=1\textwidth]{pictures/fig2.png}
  \caption{Trend of the reported attacks from the beginning of the OAD survey
    in 2008~\cite{oad20}. This comparison is only valid as a trend and not as a statistical
    measure, since the respondents are different from year to year. The dotted red line shows
    successive waves: after a relative peak of attacks, companies react and strengthen
    their cybersecurity measures, so that, the following year, the trend decreases.}
  \label{fig:2}
\end{figure}

@@

The chart in Fig.~\ref{fig:2} analyzes the trend of cyberattacks.
The figure shows that, from 2007 to 2015, in average, about 40\% of the respondents
reported cyberattacks (leftmost full red line). Since 2015, the trend increased to 45\%
of the respondents (rightmost full red line). 
In 2018, OAD reached the peak of reported attacks and, for the first time, the percent of
reported attacks became larger than that percent of companies that did not report any attack.
After the peak 
in 2018, year 2019 has a lower figure and the same occurs in the first quarter of 2020.
The trend for full 2020 will be discovered with the currently ongoing OAD survey.
This view, over several years, does not show significant changes, not even inside
the pandemic period, but shows only a slight increase in cyberattacks 
in the recent years, not necessarily attributable to the pandemic crisis. One could wonder
why the percentages in Fig.~\ref{fig:2} are lower than what the reader would have expected, 
given the numerous cyber attacks occurred in Italy. As already discussed at the beginning of 
this section, the main reason seems the size of the Italian companies.

@@

As pointed out in Fig.~\ref{fig:1}, the IT systems considered in the 2020 survey and their cybersecurity are in the medium to high range and the collected information shows a relevant 
improvement in both technical and organizational measures, in comparison to the previous surveys. Few of the considered Italian IT systems are based on data centers in 
Italy: most of the respondents' companies have medium to small IT systems partly on premise and partly outsourced, with an increasing use of cloud services.
The high number of attacks detected in 2018, and the privacy obligations related to the European
General Data Protection Regulation (GDPR), have certainly contributed to strengthening cybersecurity measures, and a further improvement 
comes from the growing use of cloud services, where high standard security measures are in place, usually. Organizational measures for cybersecurity, historically missing and neglected
in Italy, have improved in terms of definition of roles and separation of duties, of organizational policies and procedures, and of incident management. These measures often are missing
in small organizations, and in general the cybersecurity awareness and competence are low, and mainly concentrated at the top level of the public and private organizations. In Italy, there is 
still a long way to go in terms of cybersecurity continued training and awareness. A formal and bureaucratic approach to the organizational procedures is still prevalent. Once defined,
such procedures often do not find concrete application and misses periodic updates and operational tests. A significant example is the disaster recovery plans, for 
which, typically, the required alternative IT resources are neither forecast nor allocated. Periodic exercises and simulations of the 
possible disasters are not implemented. Cybersecurity management tools have limited diffusion yet, among the 2020 OAD respondents. This is particularly true for
the most advanced ones, based on artificial intelligence.
Although the use of IoT~\cite{JK21}, industrial automation, robotics systems~\cite{SGLMDXHKCZT21} and systems based on blockchain technology~\cite{RMFF21}
are expanding and can contribute to fight the Covid-19 pandemic.
In the OAD 2020 report, one finds only a few respondents involved in such technologies, which also derives from the limited portion
of their economic sector, interested in such systems: the manufacturing sectors, the logistics, the research and 
development centers and labs, the local public administration for territorial control, and so on.
%For these topics, the data that emerged from the survey are considered only as 
%specific cases that contribute to the general values on the cyberattacks, but which cannot be still considered of reference at the Italian level.
